Regularised parameter selection in the non-stationary Hawkes process

### 2017-05-01 _paper_output/chapter_sparse_hawkes.tex
more Zifying

This is analaogous to a blind deconvolution problem
\cite{stockham_blind_1975}, adaptive robust regression
\cite{donoho_automatic_1988}, or a sparse inverse problem
\cite{tibshirani_regression_1996}, or, but considerably more
challenging; While the process model superficially resembles a linear
system, there are few of the convenient results that would apply in a
linear regression problem; and as a time series problem it is difficult
to choose our penalty parameters by standard non-parametric means such
as cross validation. The proposed method uses a heuristically-motivated
degrees-of-freedom regularisation parameter selection, which we test
through simulation. The proposed method works well in practice.

@@

We are given a counting process \(N(t)\) on \(\mathbb R\)

@@

We set \(\lim_{s\rightarrow 0} N(s)\equiv 0\) and
\(t_1\equiv \inf \mathbf{t}'\); Indices \(i\leq 0\) are reserved for
point in the "history" of the process before time \(0\), which may be
generated by any arbitrary process. We will not be concerned with
evaluating the likelihood of such histories but may retain them to
control for their influence on the future of the series. For the rates
to be well-defined, however, we will assume that the increment sof the
process are finite on any finite interval.

@@

\[\lambda(t|\mathcal{F}(N(t))):=\lim_{h\to 0} \frac{\mathrm E\left(N(t,t+h)\right)}{h}.\]

@@

Such regular point processes generalise the homogeneous Poisson rate in
the sense that if we choose \(\lambda*(t)\equiv \lambda_0\) then it is
the intensity of a homogeneous Poisson rate with fixed intensity
\(\lambda_0\).

We will use the for the inter-arrival waiting times \(s_i=t_{i+1}-t_i\)
to calculate the data likelihood, after Rubin \cite{rubin_regular_1972}.

    The waiting times are exponentially distributed given the intensity,
i.e. their densities \(f_i(s)\) for any given transition are given
\[\begin{aligned}
\frac{d}{dt}P(t_{i+1}=t|\mathcal{F}(N(t)))
\end{aligned}\] fa \[\begin{aligned}
f_i(t)&\equiv f(t_{i+1}-t_i|\mathcal{F}(N(t_i)))\\
&=\Lambda(t_i,t|\mathcal{F}(N(t_i))\exp\left(-t\Lambda(t_i,t|\mathcal{F}(N(t_i))\right)
\end{aligned}\]

    It follows that the log likelihood \(\ell_i:=\log f_i\) of such a
waiting time \(s_i\) is given

\[\begin{aligned}
\ell_i(t_{i+1}-t_i) &=\log \Lambda(s_|\mathcal{F}(N(t_i)) - s_i \Lambda(s_i|\mathcal{F}(N(t_i))\\
&= \log \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds - {t_{i+1}} \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds
\end{aligned}
\]

    We may therefore write the likelihood for the data \[
\begin{aligned}
\ell(\mathbf{t}) &= \sum_{1\leq i\leq N(T)}\ell_i(t_{i+1}-t_i)\\
&= \sum_{1\leq i\leq N(T)} \left( \log \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds - {t_{i+1}} \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds\right)\\
&= \sum_{1\leq i\leq N(T)} \log \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds - \sum_{1\leq i\leq N(T)} {t_{i+1}} \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds\\
&= \sum_{1\leq i\leq N(T)} \log \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds - \sum_{1\leq i\leq N(T)} {t_{i+1}} \int_{t_i}^{t_{i+1}} \lambda(s|\mathcal{F}(N(t_i))) ds
\end{aligned}
\]

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_15_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

\[
\begin{align*}
{ \scriptstyle \frac{\partial}{\partial\omega_j} }\pi\|\boldsymbol \omega_\theta(t)\|_1 &= { \scriptstyle \frac{\partial}{\partial\omega_j} }\pi \int_{\tau_{j-1}}^{\tau_j} \left|\omega_j\right|dt\\
&=\pi\Delta\tau_j\operatorname{sgn} \omega_j\\
{ \scriptstyle \frac{\partial}{\partial\omega_j} }\lambda_\theta(t|\mathcal{F}_t) &=  \Delta\tau_j\mathbb{I}_{[\tau_{j-1},\tau_j)}(t)\\
{ \scriptstyle \frac{\partial}{\partial\omega_j} } L_\pi(\mathbf t;\theta) &=-\Delta\tau_j-\pi\Delta\tau_j\operatorname{sgn} \omega_j + \sum_{\tau_{j-1}\leq t_i\leq \tau_j} \frac{1}{\lambda_\theta(t_i)}
\end{align*}
\]

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_31_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_31_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_31_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_35_0.png}
    \end{center}
    { \hspace*{\fill} \\}

@@

Here we follow a standard "thinning" method of simulating Hawkes
processes, specialised from Algorithm 7.5.IV of Daley and Vere-Jones'
\emph{Introduction to the Theory of Point Processes Volume 1}
\cite{daley_introduction_2003}, restating Ogata's original Algorithm 2
(\cite{ogata_lewis_1981}.

Since that is a general case, here we present, informally, a simpler,
specific version for linear Hawkes processes.

For clarity, we will use the representation of the Hawkes process as an
immigrant-birth process, where the recursive "self-excited" component
corresponds to the "births". We will assume that we are given a list of
arrival times of immigrants, \(\{a_1,a_2,\dots,a_N\}\), and wish to
simulate the births arising from these immigrants. We will associate a
point and its arrival time and discuss them interchangeably where
ambiguity does not arise.

Each of these \(a_i\) gives rise to a family of descendants
\(\mathrm{descendants}(a_i)\), distributed independently of all other
families, whose combined superposition is the final Hawkes process
(\cite{hawkes_cluster_1974}. It will suffice to simulate the arrival
times of each of these families. Without loss of generality we may
consider the problem of simulating the births in a single family started
at time \(a_1=0\) and translate it.

Now the families
\(\mathrm{descendants}(t_j)\forall t_j\in \mathrm{children}(a_i)\) are
all once again identically and mutually independently distributed.
Proceeding by induction, it suffices to simulate the children of a
single point at time \(0\). The same method will suffice for its
children, children of children and so on. The final realisation can be
constructed from these simulations by appropriate concatenation and
translation of the generations of each family.

The single kernel simulation is simple, since the Poisson intensity
process is no longer conditional on the history of the sequence. As
before, the influence kernel is \(\phi\). We will assume that \(\phi\)
is a.e. continuous \(L_1\) integrable and eventually decreasing in
\(t\). We associate with \(\phi\) a "majorant", \(\phi^{**}\) such that
\[\phi^{**}(t)\geq\phi(t), \forall t>0\] and \(\phi^{**}\) is
non-increasing and integrable. As an aside, we mention that for the
unimodal influence kernels used in the current paper, such majorants are
trivial to construct.

@@

Our goal is to simulate the set \(S=\mathrm{children}(a_0)\).

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Set \(i=1, t=0\), \(S=\varnothing\)
\item
  Generate \(\Delta\sim \operatorname{Exp}(1/\phi^{**}(t))\)
\item
  Set \(t\leftarrow t+\Delta\)
\item
  If \(t\geq T\), halt.
\item
  With probability \(P=\frac{\phi(t)}{\phi^{**}(t)}\), set
  \(S\leftarrow S\cup \{t\}\).
\item
  Continue from step 2.
\end{enumerate}

@@

>>>>

This is analagous to a blind deconvolution problem
\cite{stockham_blind_1975}, adaptive robust regression
\cite{donoho_automatic_1988}, or a sparse inverse problem
\cite{tibshirani_regression_1996}, but considerably more challenging;
While the process model superficially resembles an i.i.d. linear
process, the different noise model and sequential dependence mean that
there are few of the same convenient results; We cannot easily, for
example, choose our our penalty parameters by standard cross validation.
The proposed method uses a heuristically-motivated degrees-of-freedom
regularisation parameter selection, which we test through simulation.
The proposed method works well in practice.

@@

We are given a process
\(N:\mathcal{B}(\mathbb{R})\mapsto \mathbb{N}^0\), which counts the
number of arrivals in a given time. Since we are usually concerned with
arrivals over increasing time intervals, we will usually consider \(N\)
to be indexed by time: \(N(t):=N([0,t))-N((-\infty, t))\).

@@

We set \(t_1\equiv \inf \mathbf{t}'\); Indices \(i\leq 0\) are reserved
for point in the "history" of the process before time \(0\), which may
be generated by any arbitrary process. We will not be concerned with
evaluating the likelihood of such histories but may retain them to
control for their influence on the future of the series. For the rates
to be well-defined, however, we will assume that the increments of the
process are a.s. finite on any finite interval.

@@

\[\lambda(t|\mathcal{F}(N(t))):=\lim_{h\to 0} \frac{\mathrm E\left(N([t,t+h))\right)}{h}.\]

@@

Such regular point processes generalise the homogeneous Poisson rate in
the sense that if we choose \(\lambda^*(t)\equiv \lambda_0,\) then it is
the intensity of a homogeneous Poisson rate with fixed intensity
\(\lambda_0\).

    We will use the inter-arrival waiting times \(s_i=t_{i+1}-t_i\) to
calculate the data likelihood, after Rubin \cite{rubin_regular_1972}.

From the definition of the intensity we find \[\begin{aligned}
\frac{d}{dt}P(t_{i+1}=t|\mathcal{F}(N(t)))=
\end{aligned}\]

TODO: Which bits does Z want proven? Rewrite to start from transition
probability? Probably makes this shorter. Otherwise prove equivalence of
intensity and transition prob DE?

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_13_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

\[\begin{aligned}
{ \scriptstyle \frac{\partial}{\partial\omega_j} }\pi\|\boldsymbol \omega_\theta(t)\|_1 &= { \scriptstyle \frac{\partial}{\partial\omega_j} }\pi \int_{\tau_{j-1}}^{\tau_j} \left|\omega_j\right|dt\\
&=\pi\Delta\tau_j\operatorname{sgn} \omega_j\\
{ \scriptstyle \frac{\partial}{\partial\omega_j} }\lambda_\theta(t|\mathcal{F}_t) &=  \Delta\tau_j\mathbb{I}_{[\tau_{j-1},\tau_j)}(t)\\
{ \scriptstyle \frac{\partial}{\partial\omega_j} } L_\pi(\mathbf t;\theta) &=-\Delta\tau_j-\pi\Delta\tau_j\operatorname{sgn} \omega_j + \sum_{\tau_{j-1}\leq t_i\leq \tau_j} \frac{1}{\lambda_\theta(t_i)}
\end{aligned}\]

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_29_0.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_29_1.pdf}
    \end{center}
    { \hspace*{\fill} \\}

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_29_2.pdf}
    \end{center}
    { \hspace*{\fill} \\}

@@

    \begin{center}
    \adjustimage{max size={0.9\linewidth}{0.9\paperheight}}{chapter_sparse_hawkes_files/chapter_sparse_hawkes_31_0.png}
    \end{center}
    { \hspace*{\fill} \\}

@@

@@

Here we follow a standard "thinning" method of simulating Hawkes
processes, specialised from Algorithm 7.5.IV of Daley and Vere-Jones'
\emph{Introduction to the Theory of Point Processes Volume 1}
\cite{daley_introduction_2003}, restating Ogata's original Algorithm 2
(\cite{ogata_lewis_1981}. Here we present, informally, a simpler,
specific version for linear Hawkes processes.

For clarity, we will use the representation of the Hawkes process as an
immigrant-birth process, where the recursive "self-excited" component
corresponds to the "births". We will assume that we are given a list of
arrival times of immigrants, \(A=\{a_1,a_2,\dots,a_N\}\) from a
realisation of the immigrant process. In this case, that immigrant
process is, specifically, the inhomogenous Poisson process with
unconditional intensity \(\mu(t)\). We associate a birth process \(X_i\)
with each immigrant \(a_i\). Each such process is distributed
independently of all other families, whose combined superposition is the
final Hawkes process (\cite{hawkes_cluster_1974}.

The clusters are generated as follows. For each \(a_i \in A\),
independently generate children \(a_{i1},a_{i2}\dots\) according to a
Poisson process with intensity \(\phi(t-a_i)\). Then, each child
\(a_{ij}\) of \(a_i\) again generates a Poisson process with intensity
\(\phi(t − a_{ij})\), and so on. The collection of all generated points
from all descendants forms the Hawkes process.

It suffices to know how to simulate a Poisson process with inhomogenous
rate \(\phi\) started any point \(t_i\). We will assume that \(\phi\) is
a.e. continuous \(L_1\) integrable and eventually decreasing in \(t\).
We associate with \(\phi\) a "majorant", \(\phi^{**}\) such that
\[\phi^{**}(t)\geq\phi(t), \forall t>0\] and \(\phi^{**}\) is
non-increasing and integrable. As an aside, we mention that for the
unimodal influence kernels used in the current paper, such majorants are
trivial to construct.

@@

Our goal is to simulate the set \(S=\mathrm{children}(a_\chi)\).

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Set \(i=1, t=t_i\), \(S=\varnothing\)
\item
  Generate \(\Delta\sim \operatorname{Exp}(1/\phi^{**}(t))\)
\item
  Set \(t\leftarrow t+\Delta\)
\item
  If \(t\geq T\), halt.
\item
  With probability \(P=\frac{\phi(t)}{\phi^{**}(t)}\), set
  \(S\leftarrow S\cup \{t\}\).
\item
  Continue from step 2.
\end{enumerate}

@@

    TODO
